Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
Tweak spacing of visible space char

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}

@@

%\newcommand{\QOpt}{Optician_Q}
\newcommand{\QOpt}{Optician\textsubscript{Q}}
\newcommand{\OpticianRuntime}{\textbf{Optician}}
\newcommand{\QREOptician}{\textbf{Optician\textsubscript{Q}}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

We offer three interconnected innovations to simplify the task of programming
quotient lenses.  
First, we introduce {\em quotient regular expressions} (QREs), to
easily mark aspects of data formats that are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular language and an equivalence
relation on this language.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, i.e., lenses mapping
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into one of the form $c_1 ; \ell ;
c_2^{-1}$, where $c_1$ and $c_2^{-1}$ are {\em canonizers}, $\ell$ is a bijective
transformation, and $;$ is composition\iflastminute\bcp{$c_2^{-1}$ is ugly, but I think
  we need the superscript for accuracy.  Perhaps calling them $c$ and $d$
  instead of $c_1$ and $c_2$ would be better?}\fi.  Intuitively,
the canonizers normalize data by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we leverage this normalization theorem to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs, reusing earlier work on
synthesizing plain bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the Optician benchmark suite.

@@

\emph{Bijective lenses} are amongst the most restrictive. 
A bijective lens $\ell : S \Leftrightarrow T$ specifies a bijection from
a source set $S$ to a target set $T$:
\begin{equation}\label{bijectivelenslaws}
\ell.\get \; (\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s)
= s
\end{equation}
Since the \emph{get} and \emph{put} functions form a bijection, 
no information is lost during translation from one data format to another. 
While this very strong guarantee can sometimes be useful, it is more common for 
some of the details in one or both of the data formats to be
superfluous. For instance, the number of white space characters
between data items, the capitalization of various strings, the
sequence of fields in a record, or the order of items in a list are
often irrelevant.  Also, many
data formats introduce their own ad hoc conventions about formatting details
that are taken to be inessential. In \bibtex{}, for example,
an author can be specified either as 
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''.

{\em Quotient lenses}~\cite{quotientlenses} loosen the strong restrictions of
bijective string lenses by including equivalence
relations on the source and target formats and requiring that the
round-tripping lens laws 
hold  modulo these equivalences, rather than ``on the nose.'' Quotient lenses 
make it possible to synchronize many more data formats than is possible with
plain bijective lenses\bcp{Do we justify this claim later?}. 
%
The bidirectional programming language {\em
Boomerang}~\cite{boomerang} 
implements a general class of quotient lenses.  Unfortunately, these
lenses to have two main drawbacks.  \bcp{This wording may raise a question
  in the reader's mind about whether this paper is really about quotient
  lenses in general or whether it is specifically a proposal to tweak
  Boomerang.  I.e., ``if I am not a Boomerang user, do I want to bother
  reading this paper?''  My answer would be yes, because quotient
  lenses and QREs are interesting in themselves.  But I don't think we've
  made this point very convincingly in the way we've framed the paper.}

The first shortcoming is that defining an equivalence
relation in Boomerang requires explicitly providing a {\em canonizer}, 
a pair of functions $(f,g)$ where $f$
normalizes data from the
external format by choosing canonical representations
and $g$ provides an
inverse by specifying the external form of each canonical
representation. 
The disadvantage of this approach is that writing
a quotient lens in Boomerang requires separately defining four
different data formats: two ``external'' formats which specify the data to
be transformed by the lens, and two ``internal'' formats which specify the
canonical representations for the source and target data\bcp{Plus writing
  two canonizers!}.

Our solution to this shortcoming is to introduce {\em Quotient Regular
Expressions} (QREs), which enable programmers to simultaneously specify a
regular expression and a set of canonical representatives for matching
strings.  Given an QRE, we can automatically infer the corresponding
canonization functions.  As an example, consider the following QRE for writing
author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

The second drawback with writing quotient lenses in Boomerang is that even
after specifying the four required data formats, the programmer must still
define a lens for converting between the canonical representations.
Experience shows that writing such lenses can be tedious (e.g., consider
having to rearrange the order of data items by recursively using operators
to \kw{swap} adjacent fields) and that satisfying the type checker can be
difficult because the most obvious way to express a lens runs afoul of one
or more ``ambiguity checks'' that the type checker performs in order to make
sure that the lens laws hold~\cite{optician}.

Our solution to this problem is to introduce {\em QRE lenses}, which
are a new class of quotient lens that can be automatically
synthesized, obviating the need to write lenses by hand.  A QRE lens
is like a normal lens except it maps between formats described by
QREs. We prove a normal form theorem (Theorem~\ref{normal form}) that says that
every composition of QRE lenses can be rewritten 
to be the composition of a source canonizer, a bijective lens, and a
target canonizer. This normalization property means we can synthesize
QRE lenses by extending an existing algorithm for synthesizing
bijective lenses~\cite{optician}.

@@

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}
binds \codefont{l} to a synthesized QRE lens mapping between names in the
comma-separated form describd by the QRE \codefont{comma\_name} and the
space-separated form described by the QRE \codefont{space\_name}.  The
lenses generated by our synthesis algorithm are guaranteed to translate data
back and forth between source and target QREs and to satisfy the provided
examples.  Our synthesized quotient lenses have a relatively restricted
form, where the sequential composition operator is not used and all
canonization occurs at the very edges (which greatly speeds up synthesis),
but the normalization theorem means this restriction does not compromise
expressiveness.

>>>>

\lstset{ language=Caml, basicstyle=\linespread{0.8}\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=true,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens,let},
emphstyle=\upshape\sffamily\color{dkred}, 
columns=fullflexible,
escapechar=\#,
mathescape, 
xleftmargin=1.5em,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
}
\makeatletter
     \let\lst@oldvisiblespace\lst@visiblespace
     \def\lst@visiblespace{\,\lst@oldvisiblespace\,}
\makeatother

@@

%\newcommand{\QOpt}{Optician_Q}
\newcommand{\QOpt}{QRE-enhanced Optician}
\newcommand{\OpticianRuntime}{\textbf{Optician}}
\newcommand{\QREOptician}{\textbf{Optician\textsubscript{Q}}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

Building on recent work that showed how to synthesize bijective lenses from high
level specifications, we offer three innovations to synthesize quotient lenses.
First, we introduce {\em quotient regular expressions} (QREs), that easily mark
aspects of data formats that are inessential and hence give a high-level
specification of regular languages with an equivalence relation defined on
them. Next, we introduce {\em QRE lenses}, i.e., lenses mapping between QREs.
Our key technical result is a proof that every QRE lens can be transformed into
one of the form $c_1 ; \ell ; c_2^{-1}$, where $c_1$ and $c_2^{-1}$ are {\em
canonizers}, $\ell$ is a bijective transformation, and $;$ is
composition\iflastminute\bcp{$c_2^{-1}$ is ugly, but I think we need the
superscript for accuracy.  Perhaps calling them $c$ and $d$ instead of $c_1$
and $c_2$ would be better?}\fi.  Intuitively, the canonizers normalize data by
picking a canonical representative of the relevant equivalence relations. This
theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation. Finally, we leverage this normalization theorem to {\em
synthesize} QRE lenses from a pair of QREs and example input-output pairs,
reusing earlier work on synthesizing plain bijective lenses. We have
implemented QREs and QRE lens synthesis in the bidirectional programming
language Boomerang. We evaluate the effectiveness of our approach by using it
to synthesize QRE lenses between various real-world data formats in the
Optician benchmark suite.

@@

In ~\cite{optician}, we developed a tool called {\em Optician} to synthesize the
class of {\em bijective lenses} that can be derived in the lens programming
language {\em Boomerang}\cite{boomerang}. Each lens $\ell : S \Leftrightarrow T$
that Optician can synthesized specifies a bijection from the language of a
source regular expression $S$ to a target language of a target regular
expression $T$: \begin{equation}\label{bijectivelenslaws} \ell.\get \;
(\ell.\lput \; t) = t \text{, and } \ell.\lput \; (\ell.\get \; s) = s
\end{equation}
Optician makes programming bijective lenses faster and easier, and though
many useful lenses are not bijective in nature, they do have a bijective core
surrounded by structured transformations, hence Optician can be used as a
drop-in component for synthesizing more complex lenses.

This paper applies this idea to synthesize {\em quotient lenses}
~\cite{quotientlenses}. Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\sim_S$ and $\sim_T$ defined on the source
and target data respectively:
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \sim_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \sim_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

Optician synthesizes bijective lenses from a pair $(S, T)$ of regular expression
specifying the source and target types and a set of example input-output pairs
that guide the synthesis algorithm. This presents a challenge for synthesizing
quotient lenses since a specification of the source and target formats for a
quotient lens needs to account for the equivalence relations defined on the
respective formats. Our solution is to introduce {\em Quotient Regular
Expressions} (QREs), which are regular expressions augmented with extra
syntax that enables programmers to simultaneously specify a regular expression
as well as an equivalence relation on the language of that regular expression.
Further, given a QRE, we can automatically infer a set of canonical
representatives the equivalence relation defined by the QRE as well as a
canonizing function to the canonical set of representatives.

For example, consider the following QRE for writing author names:
%
\begin{lstlisting}
let wsp_sp = project wsp $\to$ " "
let comma_name = first_name . "," . wsp_sp . last_name
\end{lstlisting}
%
In this example, \lstinline{wsp} is an existing regular expression for a
nonempty sequence of whitespace characters, and \lstinline{first_name} and
\lstinline{last_name} are existing regular expressions for first and last names.
The QRE \lstinline{wsp_sp} is a QRE with the same underlying language as
\lstinline{wsp}, but with a single canonical representative: \lstinline{" "}.
This is then used to build up \lstinline{comma_name}, a QRE with an
underlying language of two names, separated by a command and whitespace, where
the names with a single space between them are canonical.

Our second main contribution in this paper is to introduce {\em QRE lenses},
which are bijective quotient lenses that map between QREs via bijective lenses
that map between the respective canonical formats defined by the source and
target QREs. This idea is simple and natural, but begs the following question:
do we give up expressiveness if we restrict ourselves to this form? Our main
technical contribution to this question asserts that we do not; more
specifically, we will prove a normal form theorem (Theorem~\ref{normal form})
that says that every QRE lens can be rewritten to be the composition of a
source canonizer, a bijective lens, and a target canonizer. This normalization
property will enable us to (1) synthesize QRE lenses by extending the synthesis
algorithm used by Optician, and (2) prove that if there is a QRE lens that
satisfies the input specification, then this extended algorithm will return
such a lens.

@@

\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Lovelace, Ada", "Ada Lovelace")}
\end{lstlisting}
binds \codefont{l} to a synthesized QRE lens mapping between names in the
comma-separated form describd by the QRE \codefont{comma\_name} and the
space-separated form described by the QRE \codefont{space\_name}.
